A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-13 article/03_Results.Rmd
added text for an own narrative

Figure \ref{development_burial_type} and \ref{development_burial_construction} show the region-wise time series of trait occurrence. The relative proportion of each trait calculated based on ^14^C dates can be compared to the archaeological state of research:

General observations: **TODO**

`r a("Southeastern Central Europe")`^[Amount of graves in Radon-B for this region.]:
The artificial study region *Southeastern Central Europe* comprises the eastern part of Austria (Upper Austria, Lower Austria, Styria, Burgenland), the northwest of Hungary, Western Slovakia, Moravia and large parts of Bohemia (Figure \ref{map_regions} A). The density of ^14^C data on graves in Radon-B is low -- dates are available from the various regions, but large areas are without evidence. Diachronically the amount of data is consistently low and reaches a minimum in the Middle Bronze Age. While there is a constant flow of information for the variable *burial type*, the number of graves without information on *burial construction* increases in the Late Bronze Age to such an extent that no meaningful overview can be derived.  The region intersects diverse geographical and cultural areas complicating the comparison with archaeological observations. In the southeast of Central Europe, cremation and inhumation already existed side by side in the Early Bronze Age, and the local tradition remained stable in this respect also due to the expansion of the burial mound tradition in the Middle Bronze Age -- with a slight overall increase in cremation graves. The Urnfield period brought the transition to the absolute dominance of cremation burial. The development of the *burial type* variable in the Radon-B proxy shows this development correctly. The information from ^14^C data on *burial construction* correctly traces the relevance of flat burials in the Early Bronze Age but then overemphasises them in the Middle Bronze Age compared to the construction of burial mounds. In the Late Bronze Age, burials on flat burial grounds were frequent, although burial mounds of the Tumulus Culture were also further documented. This impact does not become clear in the Radon-B data -- indeed due to the small number of dates [@lubosCzechLandsAustria2013; @markovaSlovakiaHungary2013].

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, figure \ref{sed_region_matrix} shows the time series of squared euclidian distance for each region-region relationship. The majority of the curves in its upper right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200calBC is small in almost every region-region relationship. Only the Benelux region, where the share of cremation is already high in the Early Bronze Age, is excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the fact that the transition process in each region as depicted in Radon-B is different and that it took place with a time lag compared to other regions. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux region, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is also due to how correctly or incorrectly the proxy developed from radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.

>>>>

Figure \ref{development_burial_type} and \ref{development_burial_construction} show the region-wise time series of trait occurrence. If we ignore all other archaeological knowledge, we can construct the following narrative from them:

In the course of the Bronze Age in Europe there was a trend away from body burial towards cremation. Around 2200calBC cremation burials were almost unknown in Poland, southern Germany, northeastern France and northern Germany. In the northwest, in Great Britain and the Benelux area, and in the southeast, in Austria and the Czech Republic, however, there were early cremation traditions. These contexts could be discussed as areas of origin of the later omnipresent phenomenon. While cremation burials lost importance in Austria and the Czech Republic as well as in the Benelux region during the Early Bronze Age and only regained their importance in the Middle Bronze Age, their share increased steadily in England. Inhumation and cremation were balanced here for a long time. The situation was similar in southern Scandinavia, where the proportion of cremation burials was much lower until the Late Bronze Age. In Poland and northern Germany, cremation became relatively suddenly and completely established with the beginning of the Late Bronze Age. In northeastern France and northern Germany this change took place already in the Middle Bronze Age. In both contexts, inhumation continued to play an important role.

With regard to the question of grave mound construction, the picture is considerably more heterogeneous. In Austria and the Czech Republic, flat graves were the dominant form of burial until the Late Bronze Age, but mounds became more relevant from the Middle Bronze Age onwards. In Poland, flat and hill graves remained in balance until the Middle Bronze Age, when flat graves prevailed. The amount of data from southern Germany and northeastern France is extraordinarily small: If one believes the sample, then a sudden, radical change from burial in flat graves to hill graves took place in southern Germany at the beginning of the Middle Bronze Age. In northeastern France there were no flat graves in the Bronze Age. Northern Germany underwent a development from the dominance of flat graves in the Early Bronze Age, a short phase of increasing mound construction in the Middle Bronze Age followed by renewed dominance of the flat grave custom in the Late Bronze Age. A similar development can be seen in southern Scandinavia: Flat graves clearly predominated, but were increasingly -- almost completely -- replaced by hills, until burial in flat graves became more frequent again at the end of the Bronze Age. Burial mounds were dominant throughout the Benelux region, but in the Early and Late Bronze Ages flat graves also occurred to a significant extent. In England, flat graves were a short phenomenon in the Early Bronze Age, which did not recur later.

This narrative is wrong in many aspects. The following section contains a short comparison with the archaeological state of research for each region:

`r a("Southeastern Central Europe")`^[Amount of graves in Radon-B for this region.]:
The artificial circular region *Southeastern Central Europe* comprises the eastern part of Austria (Upper Austria, Lower Austria, Styria, Burgenland), the northwest of Hungary, Western Slovakia, Moravia and large parts of Bohemia (Figure \ref{map_regions} A). The density of ^14^C data on graves in Radon-B is low -- dates are available from the various regions, but large areas are without evidence. Diachronically the amount of data is consistently low and reaches a minimum in the Middle Bronze Age. While there is a constant flow of information for the variable *burial type*, the number of graves without information on *burial construction* increases in the Late Bronze Age to such an extent that no meaningful overview can be derived.  The region intersects diverse geographical and cultural areas complicating the comparison with archaeological observations. In the southeast of Central Europe, cremation and inhumation already existed side by side in the Early Bronze Age, and the local tradition remained stable in this respect also due to the expansion of the burial mound tradition in the Middle Bronze Age -- with a slight overall increase in cremation graves. The Urnfield period brought the transition to the absolute dominance of cremation burial. The development of the *burial type* variable in the Radon-B proxy shows this development correctly. The information from ^14^C data on *burial construction* correctly traces the relevance of flat burials in the Early Bronze Age but then overemphasises them in the Middle Bronze Age compared to the construction of burial mounds. In the Late Bronze Age, burials on flat burial grounds were frequent, although burial mounds of the Tumulus Culture were also further documented. This impact does not become clear in the Radon-B data -- indeed due to the small number of dates [@lubosCzechLandsAustria2013; @markovaSlovakiaHungary2013].

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, Figure \ref{sed_region_matrix} shows the time series of squared euclidian distance for each region-region relationship. The majority of the curves in its upper right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200calBC is small in almost every region-region relationship. Only the Benelux region, where the share of cremation is already high in the Early Bronze Age, is excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the fact that the transition process in each region as depicted in Radon-B is different and that it took place with a time lag compared to other regions. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux region, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of Figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is also due to how correctly or incorrectly the proxy developed from radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.
